Fractional Types

### 2013-02-05 frac.tex
more additions
typos

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
  In previous work, we developed a \emph{first-order},
  information-preserving, and reversible programming language {{langRev}}
  founded on type isomorphisms. Being restricted to first-order types limits
  the expressiveness of the language: it is not possible, for example, to
  abstract common program fragments into a higher-level combinator. In this
  paper, we introduce a higher-order extension of {{langRev}} based on the
  novel concept of \emph{fractional types} {{1/b}}. Intuitively, a value of a
  fractional type {{1/v}} represents \emph{negative} information. A
  higher-order function is modeled by a pair {{(1/v1,v2)}} with {{1/v1}}
  representing the needed argument and {{v2}} representing the
  result. Fractional values can be freely propagated and transformed but must
  ultimately --- in a complete program --- be offset by the corresponding
  amount of positive information.
\end{abstract}

@@

We are witnessing a convergence of ideas from several distinct research
communities (physics, mathematics, and computer science) towards replacing
\emph{equalities} by \emph{isomorphisms}. The combined programme originating
from these research areas has sparked a huge amount of research that unveiled
new and surprising connections between geometry, algebra, logic, and
computation (see \cite{baez2011physics} for an overview).

@@

Note that a small modification of {{lanRev}} (the addition of an
empty type {{0}} with its usual laws) would exactly capture, at the type
level, the notion of \emph{semiring} (occasionally called a \emph{rig}) where
we replace equality by isomorphism.  This can also be also compared to
\emph{bimonoidal categories}.  As is usual with with type theories, the
coherence conditions (pentagonal and triangle identity) are not explicit,
but would correspond to certain identity types being trivial.  {{lanRev}}
instead, at the type leve, models a commutative ringoid where the
multiplicative structure forms a commutative monoid, but the additive structure
is just a commutative magma.  We chose not to include {{0}} as a technical
convenience for the next section.

@@

For a given type {{b}}, the values of type {{1/b}} are of the form {{1/v}}
where {{v : b}}. Note that {{1/v}} is purely formal, as is {{1/b}}.
However, we think of a value like {{1/left ()}} as a
\emph{first-class constraint} that can only be satisfied if it is matched
with an actual value {{left ()}} or as representing the absence of some
information (i.e., \emph{negative information}), that can only be supplied by
{{left ()}}. The combinator {{eta*_b : 1 <-> 1/b * b}} represents a fission
point which creates --- out of no information --- an equal amount of negative
and positive information. Symmetrically, the combinator {{eps*_b : 1/b * b <->
1}} matches up an equal amount of negative and positive information producing 
no residual information.  The best analogy is probably that of vectors and
covectors, also know as column vectors and row vectors (respectively): in
the finite dimensional case, they are isomorphic, but they are nevertheless
quite different, being $(1,0)$ and $(0,1)$ tensors (respectively).  In
other words, since we are explicitly interested in \emph{isomorphisms}, 
we should not conflate negative-information values with positive-information
values just because they happen to be isomorphic.

@@

\begin{definition}[Denotation of Types {{ [[b]] }}]
\label{chx:def:denot}
Each type denotes a set of values as follows:
\[\begin{array}{rcl}
\llbracket 1 \rrbracket &=& \{ () \} \\
\llbracket \mathit{bool} \rrbracket &=& \{ \mathit{true}, \mathit{false} \} \\
\llbracket b_1 + b_2 \rrbracket &=& \{ \mathit{left}~v ~|~ v \leftarrow \llbracket b_1 \rrbracket \}
           \cup \{ \mathit{right}~v ~|~ v \leftarrow \llbracket b_2 \rrbracket \} \\
\llbracket b_1 \times b_2 \rrbracket &=& \{ (v_1,v_2) ~|~ v_1 \leftarrow \llbracket b_1 \rrbracket, 
           v_2 \leftarrow \llbracket b_2 \rrbracket \} \\
\llbracket 1/b \rrbracket &=& \{ 1/v ~|~ v \leftarrow \llbracket b \rrbracket \} 
\end{array}\]
\end{definition}

The semantics of {{langRevTF}} is given using two mutually recursive
interpreters: one going forward and one going backwards. Unlike the case for
{{langRev}}, each combinator denotes a \emph{relation} and hence the result
of applying a combinator to a value generally returns a \emph{set} of
values.

@@

The language {{langRevT}} is still logically reversible.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
  In previous work, we developed a \emph{first-order},
  information-preserving, and reversible programming language {{langRev}}
  founded on type isomorphisms. Being restricted to first-order types limits
  the expressiveness of the language: it is not possible, for example, to
  abstract common program fragments into a higher-level combinator. In this
  paper, we introduce a higher-order extension of {{langRev}} based on the
  novel concept of \emph{fractional types} {{1/b}}. Intuitively, a value of a
  fractional type {{1/v}} represents \emph{negative} information. A
  higher-order function is modeled by a pair {{(1/v1,v2)}} with {{1/v1}}
  representing the needed argument and {{v2}} representing the
  result. Fractional values are first-class: they can be freely propagated
  and transformed but must ultimately --- in a complete program --- be offset
  by the corresponding amount of positive information.
\end{abstract}

@@

We are witnessing a convergence of ideas from several distinct research
communities (physics, mathematics, and computer science) towards replacing
\emph{equalities} by \emph{isomorphisms}. The combined programme originating
from these research areas has sparked a huge amount of research that unveiled
new and surprising connections between geometry, algebra, logic, and
computation (see~\cite{baez2011physics} for an overview).

@@

Note that a small modification of {{langRev}} (the addition of an
empty type {{0}} with its usual laws) would exactly capture, at the type
level, the notion of \emph{semiring} (occasionally called a \emph{rig}) where
we replace equality by isomorphism.  This can also be also compared to
\emph{bimonoidal categories}.  As is usual with with type theories, the
coherence conditions (pentagonal and triangle identity) are not explicit,
but would correspond to certain identity types being trivial.  {{langRev}}
instead, at the type level, models a commutative ringoid where the
multiplicative structure forms a commutative monoid, but the additive structure
is just a commutative magma.  We chose not to include {{0}} as a technical
convenience for the next section.

@@

For a given type {{b}}, the values of type {{1/b}} are of the form {{1/v}}
where {{v : b}}. Note that {{1/v}} is purely formal, as is {{1/b}}.
However, we think of a value like {{1/left ()}} as a
\emph{first-class constraint} that can only be satisfied if it is matched
with an actual value {{left ()}} or as representing the absence of some
information (i.e., \emph{negative information}), that can only be supplied by
{{left ()}}. The combinator {{eta*_b : 1 <-> 1/b * b}} represents a fission
point which creates --- out of no information --- an equal amount of negative
and positive information. Symmetrically, the combinator 
{{eps*_b : 1/b * b <-> 1}} matches up an equal amount of negative and positive
information producing no residual information.  The best analogy is probably
that of vectors and
co-vectors, also know as column vectors and row vectors (respectively): in
the finite dimensional case, they are isomorphic, but they are nevertheless
quite different, being $(1,0)$ and $(0,1)$ tensors (respectively).  In
other words, since we are explicitly interested in \emph{isomorphisms}, 
we should not conflate negative-information values with positive-information
values just because they happen to be isomorphic.

@@

\begin{definition}[Denotation of Value Types {{ [[b]] }}]
\label{chx:def:denot}
Each type denotes a set of values as follows:
\[\begin{array}{rcl}
\llbracket 1 \rrbracket &=& \{ () \} \\
\llbracket \mathit{bool} \rrbracket &=& \{ \mathit{true}, \mathit{false} \} \\
\llbracket b_1 + b_2 \rrbracket &=& \{ \mathit{left}~v ~|~ v \leftarrow \llbracket b_1 \rrbracket \}
           \cup \{ \mathit{right}~v ~|~ v \leftarrow \llbracket b_2 \rrbracket \} \\
\llbracket b_1 \times b_2 \rrbracket &=& \{ (v_1,v_2) ~|~ v_1 \leftarrow \llbracket b_1 \rrbracket, 
           v_2 \leftarrow \llbracket b_2 \rrbracket \} \\
\llbracket 1/b \rrbracket &=& \{ 1/v ~|~ v \leftarrow \llbracket b \rrbracket \} 
\end{array}\]
\end{definition}

The operational semantics of {{langRevTF}} is given using two mutually
recursive interpreters: one going forward and one going backwards. Unlike the
case for {{langRev}}, each combinator denotes a \emph{relation} and hence the
result of applying a combinator to a value generally returns a \emph{set} of
values.

@@

The language {{langRevTF}} is still logically reversible.
